The Key to a Data Parallel Compiler

### 2015-11-11 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
Complete the other passes section.

\documentclass[pldi]{sigplanconf-pldi15}
% \usepackage[utf8]{inputenc}
\usepackage{fontspec}
\setmainfont[Ligatures=TeX]{STIX}
\setmonofont{APL385 Unicode}
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

@@

We will use these keys to perform computation over the AST and accomplish our tasks of function lifting and 
expression flattening in the next section, but we make a final note here, that we can imagine many other 
operations which might be used throughout the compiler when we care about how nodes relate to each 
other. The use of operators like reduction, scan, inner and outer products allow us to obtain the necessary 
information from the node coordinate matrix. Sometimes these results might not be obvious, but the 
example of the “belongs to” relationship given above demonstrates a pattern that arises repeatedly 
throughout our data parallel compiler, and is therefore particularly useful to understand. 
%[cf]
%[of]:Computing with Node Coordinates
\section{Computing with Node Coordinates}

Given the above keys and grouping information, we now have all the information we need to do 
function lifting and expression flattening, but we lack a suitable operator to complete the task. 
Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
analytics operator, represents just such a useful tree transformation tool.

In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
usually the Module boundary node which contains the entire set of functions and values in the module.
%[cf]
%[of]:Function Lifting
\subsection{Function Lifting}

@@

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[cf]
%[of]:Scaling up, Considerations and Thoughts
\section{Scaling up, Considerations and Thoughts}

Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
Besides brevity and directness, the complexity of each operation and the composition of operations is 
well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
conceivable then, that the compiler could produce not only an executable program from code written 
in this style, but also a complexity bound on its execution. 
%[cf]
%[of]:Related Work
\section{Related Work}

>>>>

\documentclass[pldi]{sigplanconf-pldi15}
% \usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{fontspec}
\setmainfont[Ligatures=TeX]{STIX}
\setmonofont{APL385 Unicode}
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

@@

We will use these keys to perform computation over the AST and accomplish our tasks of function lifting and 
expression flattening in the next section, but we make a final note here, that we can imagine many other 
operations which might be used throughout the compiler when we care about how nodes relate to each 
other. The use of operators like reduction, scan, inner and outer products allow us to obtain the necessary 
information from the node coordinate matrix. Sometimes these results might not be obvious, but the 
example of the “belongs to” relationship given above demonstrates a pattern that arises repeatedly 
throughout our data parallel compiler, and is therefore particularly useful to understand. 
%[cf]

%[cf]
%[cf]
%[of]:Case Studies
\section{Case Studies}

%[of]:Function Lifting
\subsection{Function Lifting}

@@

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[of]:Lexical Scoping
\subsection{Lexical Scoping}
%[cf]
%[cf]
%[of]:Table of Compiler Passes
\renewcommand{\arraystretch}{1.5}
\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\textbf{Pass} & \textbf{Description} & \textbf{Core Patterns} 
 & \textbf{Coordinate?} \\ \hline
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{table:passes}
\end{table*}
%[cf]
%[of]:Other Passes
\section{Other Passes}

Table \ref{table:passes} contains a selection of passes from the Co-dfns compiler 
and breaks down their features based on how they traverse the tree and whether 
or not they use the Key operator or Node Coordinates. They are oriented with 
front-end passes higher in the table and passes further down the compiler 
chain at the bottom of the table. These passes give a good idea of how the Key 
operator plays a part in the complete compiler as well as how Node Coordinates 
help to deal with complex sub-tree selections.

We make a few notes on the table and the analysis. The Key operator can be used 
with or without the node coordinates structure. In cases where the AST is 
sufficiently flat and arranged in a sufficiently convenient order, the grouping 
operations are very simple. In these cases, the Key operator amounts to a 
map operation over some selected sub-trees selected at a specific depth. These 
occur later in the compiler where the reader will note appearences of the Key 
operator without the use of Node Coordinates.

Likewise, node coordinates find their place early on in the compiler where it 
is necessary to deal with the more highly nested AST. They are useful for identifying 
this information regardless of whether the Key operator is used to handle the 
grouping or not. Thus passes like ``Drop Unreachable Code'' that remove whole 
sub-trees from the AST may use the Node Coordinates to do the grouping once 
the correct parent node has been identified.

In short, the goal of the structure and design of the compiler passes given 
in table \ref{table:passes} is to utilize Node Coordinates to remove as quickly as possible 
the nested structure of the AST into a flatter form that does not require complex 
analysis to traverse. 
%[cf]
%[of]:The Co-dfns Compiler
\section{The Co-dfns Compiler}
%[cf]
%[of]:Performance
\section{Performance}
%[cf]
%[of]:Future Work
\section{Future Work}
%[cf]
%[of]:Related Work
\section{Related Work}
